Convergent Evolution During Local Adaptation to Patchy Landscapes

### 2015-09-05 patchy-review-responses-second-round.tex

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We would like to submit
a revised version of our manuscript,
``Convergent Evolution During Local Adaptation to Patchy Landscapes'',
for consideration for publication in PLoS Genetics.
We have dealt with all the reviewer's comments
(which were useful),
as detailed below,
and have further edited the text for readability.

The main change was that we found where
the missing factor of area pointed out by the reviewers
had slipped through the cracks in our argument;
this required fairly minor re-writing,
and improved somewhat the fit of the theory to simulations.

We also turned up a minor bug in the discussion of conditioned branching processes:
the exact picture is not quite so simple as was previously implied,
but the approximate picture is still the same --
see around \llname{ll:branching_process_fix}.

Finally, we have further endeavored to bring the intuitive picture to the fore
and otherwise cut down on needless words.
In particular, we have 
re-written parts of the section \secref{ss:heuristics},
substantially shortened the discussion of branching processes
in the section \secref{ss:migrant_math},
and cut entirely the trailing section on ``outflux of migrant families''.

We hope you find the paper as improved as we do.

\vspace{1in}

@@

\noindent \hspace{4em}
\begin{minipage}{3in}
\noindent
{\bf Sincerely,}

\vskip 2em

{\bf Peter Ralph, and\\
Graham Coop.}\\
\end{minipage}

\pagebreak

\textbf{Response to Letter from the Guest Editor:}

\begin{quote}
    \it
    Both reviewer 1 and reviewer 2 still question your main result presented in Eq. 11 and I do agree with them that there is definitely something wrong with the dimensions. 
    % This was already pointed in the first round of reviews and both reviewers feel that this issue has not been satisfactorily addressed. 
\end{quote}

\reply
Thanks to the reviewers for sticking to this;
on revision we uncovered the error in the derivation of equation \eqref{eqn:migrate},
that supplied the missing factor of area.
The error occurred in the transition between section \secref{ss:heuristics} 
and section \secref{ss:hitting_occupation}
because in the former section we had tried to keep things simple
by working with densities in gestalt equations \eqref{eqn:gestalt_q} and \eqref{eqn:gestalt_migrate},
but then (more correctly) treated the shape in the latter section;
effectively we were treating the patch as very small in one place
and large in another.
We've rewritten the argument (section \secref{ss:heuristics})
to deal explicitly with the integral over the whole patch,
which also necessitated adding a new (short) section (\secref{apx:qS})
that deals with the geometry of the patch.
Section \secref{ss:hitting_occupation} required very little change.

\begin{quote}
    \it
    Both reviewers also argued in the first round that the derivations in the paper are overly complicated and that the problem can in fact be intuitively understood via much simpler arguments (both came up with very similar approaches independently). 
    % You stated in your replies that you prefer to stick with your derivation because you consider it more solid. 
    % However, I am not even fully convinced that your Eq. 11 is more correct than the reviewers' intuitive results. 
    [\ldots] I also think that presenting and discussing the intuitive results really would be important for the paper, as originally suggested by reviewers 1 and 2. 
\end{quote}

\reply
We agree; the main argument around these results (in section \secref{ss:heuristics})
we think follows the arguments of the reviewers closely
(and arrives at the same answer; see below).
We also still provide a more detailed argument towards the end of the paper;
in our view, this gives the best of both worlds.

% \begin{quote}
%     Reviewer 1 points out that there are quite substantial deviations in your numerical results (Fig. 4) and reviewer 2 stresses that the 1D simulations won't even be able to catch the potential problems associated with the boundary of the target area. Clearly, something is not yet fully understood here and I don't see how the paper can be accepted without having settled these issues.
% \end{quote}

% \begin{quote}
%     I also think that presenting and discussing the intuitive results really would be important for the paper, as originally suggested by reviewers 1 and 2. There is a tendency in population genetics to derive complicated ``exact'' results that are nonetheless of only limited usefulness in reality because it's unclear how realistic the underlying models are in the first place. I won't hide here that I strongly prefer a heuristic approach that reveals what's actually going on, over a lengthy technical calculation that may be formally more correct, yet it's unclear whether this really matters in reality. I feel the majority of the audience of PLoS Genetics is probably with me on this.
% \end{quote}

@@

%%%%%%
\point{}{
    The issue of dimensions and geometry of the target patch is the main difficulty I have with the paper. I just don't see how your equation \eqref{eqn:migrate} can be correct. It has dimensions of 1/time/area when it should be 1/time. Somehow the target area needs to play a role. Reviewer 2 has raised similar issues.
}

\reply
Now that we've found the problem in the derivation,
we heartily agree;
the missing factor was, indeed, a multiplicative factor of area.
(See above for more on this.)

%%%%%%
\point{}{
Regarding the simulations presented in Fig 4, I don't think they strongly support the functional form of $\migrate$. Points with very similar observed migration time have predicted migration times that differ by a factor of a 100 (the more extreme are outside the validity of the approximations, but still I don't think this figure is strong evidence that the math is correct.)
}

\reply
We think the main issue here was inclusion of points outside the validity of the approximations,
since of the parameters falling in the range of validity discussed in \secref{ss:assumptions},
the largest discrepancy was a factor of 7, not 100.
We've re-made this figure only using parameter combinations that fall within the range of validity
($R > \sigma/\sqrt{s_m}$, patch width larger than $\sigma/\sqrt{s_m} \atan(\sqrt{s_m/s_b})$, and $2 s_m \sigma N > 1$);
%         ( R*sqrt(abs(sm))/sigma > 1 ) & 
%         ( size2 > (sigma/sqrt(abs(sm)))*atan(sqrt(abs(sm/gb))) ) &
%         ( abs(sm) * 2 * sigma * N > 1 )
the figures that include all simulation results are still included in the supplement
(and hopefully will inspire theory that is applicable in the other regions).

@@

%%%%%%
\point{eq.~56 is wrong}{
    as easily seen with $\Theta=0$ \ldots
    so eq.~56 has an extra 2, resulting from a wrong expansion of $\sqrt{1+x}$,
    and so eq.~57--61 are also wrong.
}

\reply
Thanks for catching that mistake.  
% We removed the extra factor of 2 (section~\secref{apx:outflux}).
However, in the interest of making the paper less complex, we decided to remove that section.

@@

%%%%%%
\point{time in transit}{
    One can obtain these results by a simple saddle point approximation:
    \begin{align*}
        \P\{ \text{getting to $x$ in time $t$} \} &= \P\{\text{diffusion to $x$ in time $t$}\} \P\{\text{surviving}\} \\
                                                  &= \frac{1}{(2\pi\sigma^2)^{d/2}} e^{-\frac{x^2}{2\sigma^2 t}-st}
    \end{align*}
    the exponent is minimal when
    \[
        \frac{x^2}{2\sigma^2 t^2} = s \quad \Rightarrow \quad t^* = \frac{x}{\sigma\sqrt{2s}},
    \]
    this gives your eq.~17.
    The variance is simply the inverse of the second derivative of the exponent
    \[
        \frac{1}{\var[\tau]} = \frac{x^2}{\sigma^2 (t^*)^3} = \frac{\sigma}{x(2s)^{3/2}} .
    \]
}

\reply
This is a nice point,
and we've added a note to this effect \revref.
We also rewrote this section somewhat to flow more naturally:
since $\P\{L>\ell\}=\E[e^{-\ell\tau}]$ is more properly the quantity we're interested in,
we start with it.

%%%%%%
\point{now to the thornier issues}{
    \begin{itemize}
        \item {} [with $\ell = \sigma/\sqrt{2s}$, $r$ the radius of the patch, $R$ the distance between patches, and $\ell \ll r \ll R$]
        \item the average density is $q(x) \sim e^{-x/\ell}$
        \item but fluctuates far from the source as [families of size $K$, with local density $K/\ell^d$]
        \item families of size $1/s$ live for $1/s$ generations and hence contribute $1/s^2$ to the time-averaged density $q(x)$
        \item \[ q(x) = \lambda_\text{arr}(x) / s^2 \ell^d, \]
            where $\lambda_\text{arr}(x)$ is the rate at which families arrive that overlap with $x$
        \item \[ \lambda_\text{arr}(x) = q(x) s^2 \ell^d \]
            and
            \[ \Lambda_\text{arr}(x) = \lambda(x) \ell^d = q(x) s^2 \]
            would be an ``arrival density''
        \item to get at $\migrate$, arrival needs to be integrated over the relevant area. In $d=1$ this is simply a stretch of length $\ell$,
            in $d=2$ it should be $\sqrt{\ell^3 r}$,
            hence one should get
            \[ \migrate \sim \rho q(R) s^2 \ell^d \sqrt{r/\ell} (1-e^{-p_e/s}) \]
        \item result does depend on target geometry
    \end{itemize}
}

\reply
As discussed above, we fixed the bug in our argument here.
Now, \eqref{eqn:migrate} agrees with the reviewer's:
Both have $\rho q(R)$.
We have $A'$ for the area of the nearest $\ell=\sigma/\sqrt{2s_m}$ portion of the patch;
this corresponds above to $\ell^{d} \sqrt{r/\ell}$
since (see figure in reviewer report) assuming a circular patch, the transverse dimension of $A'$ is $\sqrt{r\ell}$.
The remaining term is $\E[1-(1-p_e)^K]/(\E[K]/2s_m)$;
we approximate this by $s_m \min(s_m,p_e)$; 
above it is $s^2 (1-e^{-p_e/s})$;
the two agree to first order (and they are both first order approximations).
The result does, indeed, depend on target geometry;
we've added a subfigure to figure \ref{fig:branching_decomp} to illustrate this point.

@@

%%%%%%
\point{}{
I still think that the main result (11) is dimensionally inconsistent -- it needs to be integrated over the boundary layer of the patch. In their response to my original review, the authors just referred to their response to a similar point by reviewer 1, but I don't think that this addresses my concern -- reviewer 1 was talking about integrating over a different area, the one occupied by a family, which I agree is not correct. (Note that the 1D simulations aren't going to be able to catch this issue, since you can't play with the length of the boundary.)
}

\reply
As discussed above, we fixed this.
And, apologies for not catching on in the first round of reviews.

%%%%%%
\point{}{
I do think that reviewer 1 raises a good point though. Looking back through the manuscript, I can't find where the authors show that the family is localized within a length scale $\sim \sigma/\sqrt{s_m} \ll A^{1/d}$. This is important, since it's needed to justify the approximation that the whole family arrives at the patch together. This seems like a basic fact about subcritical branching random walks/Brownian motions, so there's probably a formal proof of it somewhere in the literature, but I think that it's straightforward to give an intuitive explanation.
}

\reply
Good point; this works out nicely: \revref.
We have looked a fair bit for a formal proof in the literature and not turned it up.

@@

%%%%%%
\point{}{
Speaking of the literature, what's known about hitting probabilities for subcritical branching random walks/Brownian motions?
}

\reply
In general, we could write down differential equations
that tell us what we want to know.
There are at least two approaches: 
ignore the old patch, and simply ask for the probability of establishment;
or include the old patch, and ask about the transform of the occupation time in the new patch,
$\E[\exp(-\int_0^t \langle Z_s, p_e \rangle ds)]$.
For instance,
suppose $Z$ is a continuous-time Markovian branching process
whose spatial motion has generator $A$,
that when at location $x$, branches at rate $r(x)$,
producing a number of 
offspring whose distribution has generating function $\Psi_x(u)=\E^x[u^W]$.
Then if $f(x)$ denotes the probability of nonextinction
of a process beginning with a single individual at $x$,
\begin{align*}
    Af(x) + r(x) (\Psi_x(f(x))-f(x)) = 0 .
\end{align*}
This is the approach taken by \citep{barton1987establishment};
we chose not to go this route because thinking explicitly about the decomposition into families
seems more interesting and useful.
There is a large and diverse literature on branching Brownian motion,
but the majority of it is devoted to critical or supercritical Brownian motion.
We haven't been able to dig up anything that gives the sort of asymptotics that we need;
but it may well be out there; pointers are welcome.

@@

%%%%%%
\point{}{
I still think that the derivation of the main result is too complicated, and could be made much simpler along the lines of the suggestion in my previous review. The extra complications introduced in the current version do not actually make the result any more rigorous -- they just take a less direct route to introducing the same approximations.
}

\reply
We have gone back over the section surrounding \eqref{eqn:migrate},
aiming for maximal simplicity along these lines.
It's true, that in the case that $p_e < s_m$ we get the same result
as what is obtained ignoring correlations in arrivals
(e.g., by the Fick's law argument of the previous review).
However, we do go beyond these approximations, since we include the case $p_e > s_m$,
where family structure matters --
so, we think we've addressed this,
and hope we're not missing further simplifications.

@@

%%%%%%
\point{}{
I don't totally understand the simulation methods. The authors say that individuals first chose whether or not to reproduce with probability depending on their fitness, and then produced a Poisson-distributed number of offspring. So the selection is all in whether or not individuals reproduce, not in the number of offspring conditioned on reproduction? What exactly was the function for the probability of reproduction, and what was the mean of the offspring distribution? I'm just wondering if there's a detail here that would account for the discrepancy between the analysis and the simulations. Also, it would be helpful to say exactly what the value of sigma is in the chosen migration model.
}

\reply
Whoops -- we had erroneously left in that statement that the number of offspring each step was Poisson;
in fact, the number of offspring produced in a generation is either 0 or 1
(but, individuals may live for many generations).
The exact function is given in the next paragraph \revref,
and as detailed there, we don't naively assume
that whatever we labeled \texttt{sm} in the code
is actually what we mean by $s_m$ in the theory.
We've also added a note about what $\sigma$ is \llname{rr:sim_details}.

@@

%%%%%%
\point{}{
I'm not sure that the description of the valid parameter range in "A review, and a test of the assumptions" is entirely correct. For the mutation results, it's fine if $s_m$ and $s_p$ have very different magnitudes, as long as the patches are sufficiently big/migration is sufficiently weak. What precisely is the condition here?
}

\reply
Good point; we've been more precise about these conditions. \revref

%%%%%%
\point{}{
In the Applications, I understand that there's still a lot to learn about this system, but the authors need to at least say how far apart the actual lava flow patches are, so we have a sense of what their estimate of "tens to a few hundred kilometers" means for mutation vs migration.
}

\reply
That's a trickier question than it seems --
in particular, there doesn't exist a map of rock color
(just rock type, and the map to color is not one-to-one).
This is further complicated by the fact the mice live on rocky outcrops (only),
not on sandy or otherwise flat areas.
After consulting with Michael Nachman,
we've added a rough idea the relevant scales \revref.

@@

%%%%%%
\point{}{
In their response to my comments, the authors said that they did not see why the assumption $R \gg w$ was necessary. I think that it is needed in order to talk sensibly about a single "distance to a patch" $R$: if $w \gtrsim R$, then generally some parts of the patch will be much closer than others, right?
}

\reply
Ah: now we see what the misunderstanding was.
By "distance to a patch" we mean the length of the shortest path beginning in one and ending in the other.
We've clarified this \revref.
Given this, and the discussion of how to calculate $A'$,
we still don't think the assumption is necessary.

@@

%%%%%%
\point{}{
I still think that including any sub-leading terms in the length of the shared haplotype is falsely precise.
}

\reply
We would agree, but it's not certain which of the various terms 
in \eqref{eqn:haplotype_length} and the subsequent equation for $\var[L]$
are the leading ones,
depending on how $s_m$ compares to $\sigma/R$.

### 2015-09-05 patchy-selection-paper.tex
finished responses

%%%%%
% Responses to reviews:
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\iflatexml\reviewresponsesfalse\fi  % they don't work so well in html
% \newcommand{\responsefile}{pbio-reviews-19sept12-responses.tex}  % name of the review reponses file
\newcommand{\responsefile}{patchy-review-responses-second-round.tex}  % name of the review reponses file

@@

\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\Exp}{Exp}
\DeclareMathOperator{\sn}{sn}
\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\atan}{atan}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\one}{\mathbf{1}}
\newcommand{\calE}{\mathcal{E}}
\newcommand{\calC}{\mathcal{C}}
\newcommand{\dconv}{\xrightarrow{d}}
\newcommand{\deq}{\stackrel{\scriptscriptstyle{d}}{=}}
\newcommand{\given}{\,\mid\,}
\newcommand{\st}{\,\colon\,}

@@

Now suppose that there are two patches separated by distance $R$
(i.e., the shortest distance between the two is $R$). \revpoint{2}{9}
If the $B$ allele has arisen and become established in the first patch, but has not yet appeared in the second,
we would like to know the time scale on which copies of $B$ move between patches by migration
(supposing that no $B$ allele arises independently by mutation in the meantime).
To determine this, 
we study the fine-scale genealogy of alleles that transit between patches,
obtaining along the way other useful information about the genealogy of $B$ alleles. 
Doing this we arrive at equation \eqref{eqn:migrate} for the rate at which 
an allele established in one patch spreads to a neighboring one by migration.

@@

To understand the process of adaptation by migration,
we therefore need to look more closely at the rare families of $B$ alleles 
far from the original patch.
It is reasonable to model each such family as a branching process,
i.e., assuming that each individual
migrates, then reproduces, independently of all other family members.
Migration takes the individual in a random direction
for a random distance whose mean square is $\sigma^2$,
and if the individual arrives to a patch where the $B$ allele is favored, 
we drop the individual from the family.
Since the $B$ allele is uniformly deleterious elsewhere,
we can decouple reproduction and migration
by first forming a tree in which each individual has a random number of offspring,
then recursively assigning spatial locations to each individual in the tree,
and finally pruning any branches that happen to wander into a patch.
By our definition of fitness, 
in this tree the mean number of offspring per individual is $e^{-s_m} < 1$.
This pruning removes both branches leading back into the already established patch (which are irrelevant)
and branches leading into the new patch (we will need to count these).
Since we are concerned with the rare migrant families that transit quickly between patches, 
back migrants to the already established patch will contribute little, since they are going the wrong direction;
so to simplify the analysis we ignore this detail.  % but could add this detail in an appendix

@@

This suggests that families can be well-described by motion of a central ``trunk'',
which dies at a random time whose distribution is given by $k_e(t)$,
and conditional on survival of this trunk, the family is composed of side ``branches'' off of this trunk,
as depicted in Figure~\ref{fig:branching_decomp}.
We can understand this quite concretely, by a method described rigorously by \citet{geiger1999elementary}.
% (see also \citet{chauvin1991growing}). 
% (lyons1995conceptual deals with conditioned on eventual nonextinction)
When we condition on survival until $t$, we condition on existence of
at least one trunk lineage surviving from time $0$ to time $t$ (the long red
lineage in Figure \ref{fig:branching_decomp}).
Given this lineage, nothing distinguishes any of the other reproduction events --
each individual along the trunk lineage must give birth to at least one offspring that survives until time $t$,
and all other reproduction events are unconditioned.
In other words, the genealogy of a family that survives until $t$
can be decomposed into a trunk that lasts until $t$
and that sprouts independent branches that may or may not survive until $t$.
Specifically, if we write $Z_t$ for the number of offspring in the branching process alive at time $t$,
% (so $\P\{Z_t = 0\} = k_e(t)$),
then this gives the following decomposition:
\begin{align} \label{eqn:Zt_decomposition}
  \left( Z_t \; \vert \; Z_t>0 \right) \deq 1 + \sum_{s=0}^t \sum_{k=1}^{W_s-1} Z^{(s,k)}_{t-s},
\end{align}
where each $Z^{(s,k)}$ is an independent copy of $Z$ 
that arose from the main branch in generation $s$,
and each $W_s$ is independent with distribution 
%% equation (2.3) in Geiger:
%% $\P\{ W_s = k\} = \sum_{j=k}^\infty \P\{ W = j \} k_e(t-s-1)^{j-k} (1-k_e(t-s-1))/(1-k_e(t-s))$, 
%% and equation (2.5) in Geiger:
that tends to $\P\{W_s=k\} \simeq \P\{W\ge k\}/\E[W]$ as $s \to \infty$,
where $W$ is the unconditioned offspring number distribution.  \linelabel{ll:branching_process_fix}
In particular, $\E[W_s-1] \simeq \E[W(W-1)]/\E[W]$;
if $W$ is Poisson then this is equal to $e^{-s_m}$.
Since each subfamily $Z^{(s,k)}$ is a branching process with mean offspring number less than one,
most of these die out quickly,
and so $Z_t$ is composed of a cluster of recently split subfamilies.
If we take the expectation of \eqref{eqn:Zt_decomposition},
since $\E[Z_t] = e^{-s_m t}$,
we get the limiting mean family size conditioned on nonextinction:
\begin{align} \label{eqn:mean_K}
    \E[K] 
    = 1 + \sum_{s=0}^\infty \E[W_s-1] e^{-s_m t}
    \approx \E[W(W-1)] / ( \E[W] s_m  )
\end{align}

@@

\begin{figure}[ht!!]
  \begin{center}
    \includegraphics{branching-concept}
  \end{center}
\caption{Cartoon of the decomposition described in the text. 
Migrants families that get further than $r_0$ from the already-established patch (grey circle)
are each composed of a ``trunk'' lineage (in red), whose spatial motion we follow,
embellished by transient branches.
For the depicted long-lived family, these are further distinguished by those still alive at time $t$ (in black)
and those that did not survive until $t$ (in grey).
Figure~\ref{fig:sim_snapshots} has a depiction of such a family, from a simulation. 
Note that the spatial motions of lineages are independent random walks
(and hence should be much wigglier than in this cartoon, which confounds space and time).
} \label{fig:branching_decomp}
\end{figure}

@@

The remaining term is $q(S)$, the integral of equation \eqref{eqn:eqfreq} over the patch.
Since the patches are at distance $R$,
if the new patch has width smaller than $\sigma/\sqrt{2s_m}$, 
then $q(S)$ is approximately $q(R)$ multiplied by the area of the new patch.
If $S$ is larger, then we should use the area of the closest strip
of width $\sigma/\sqrt{2s_m}$ to the original patch;
see section \ref{apx:qS} for more details.
In either case, denote this area $A'$;
for numerical examples we compute $q(S)$ directly.

@@

To keep the expressions simple, in the remainder of this section we only compute quantities for one dimension.
By Bayes' rule, 
\begin{equation} \label{eqn:haplen_cdf}
\E[e^{-\ell \tau}|\tau<\infty]  = \exp\left\{{-\frac{R}{\sigma}\left(\sqrt{2(\ell+s_m)} - \sqrt{2s_m}\right)}\right\} .
\end{equation}
This can be differentiated to find that the expected transit time is
\begin{equation} 
  \E[\tau \given \tau<\infty ] = (R\sqrt{2s_m}/\sigma)\times 1/(2s_m) \label{eqn:mean_tau}
\end{equation}
and that $\var[\tau \given \tau<\infty ] = (R\sqrt{2s_m}/\sigma) \times 1/(2s_m)^2$.

@@

Coat color in the rock pocket mouse
\emph{Chaetodipus intermedius} is a classic example of local
adaptation 
\citep{benson1933concealing,DiceBlossom1937}.
These mice live on rocky outcrops throughout the southwestern United States and northern Mexico, 
and generally have a light pelage similar to the predominant rock color.
However, in some regions these mice live on darker substrates (e.g., old lava flows),
and in these patches have much darker pigmentation, 
presumably to avoid visual predators.
Some of the largest patches of dark rock range from 10km to 100km wide
and lie 50--400km from each other,
and dark-colored populations of \emph{C.\ intermedius} have been found on many of these.
(However, patches of all sizes occur across all scales in a heterogeneous manner.)
\revpoint{5}{7}
\citet{nachman2003genetic} demonstrated that on one of these flows (Pinacate), 
much of the change in coloration is due to an allele at the MC1R locus.
This dark allele differs from the light allele by four amino acid changes, 
and has a dominant or partially dominant effect depending on the measure of coat color.
The Pinacate allele is not present in a number of other populations with dark pelage,
suggesting these populations have adapted in parallel \citep{nachman2003genetic,hoekstra2003different}.
However, \citet{hoekstra2004local} reasoned that, elsewhere in the range, 
multiple close dark outcrops may share a dark phenotype whose genetic basis has been spread by migration
despite intervening light habitat.

@@

In the developement above
we need to approximate the integral of equation \eqref{eqn:eqfreq}
across the area occupied by the new patch,
$q(S) = \int_S q(x) dx$.
The precise answer depends on the shape and orientation of the patches;
but we aim for a usable approximation,
primarily in terms of the shortest distance from the old patch to the new patch, denoted $R$,
and assuming $R$ is large enough we can take equation \eqref{eqn:eqfreq} as an equality.

>>>>

%%%%%
% Responses to reviews:
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\iflatexml\reviewresponsesfalse\fi  % they don't work so well in html
% \newcommand{\responsefile}{pbio-reviews-19sept12-responses.tex}  % name of the review responses file
\newcommand{\responsefile}{patchy-review-responses-second-round.tex}  % name of the review responses file

@@

\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\Exp}{Exp}
\DeclareMathOperator{\sn}{sn}
\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\atan}{atan}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\one}{\mathbf{1}}
\newcommand{\calE}{\mathcal{E}}
\newcommand{\calC}{\mathcal{C}}
\newcommand{\dconv}{\xrightarrow{d}}
\newcommand{\deq}{\stackrel{\scriptscriptstyle{d}}{=}}
\newcommand{\given}{\mid}
\newcommand{\st}{\,\colon\,}

@@

Now suppose that there are two patches separated by distance $R$
(i.e., the shortest distance between the two is $R$). \revpoint{2}{8}
If the $B$ allele has arisen and become established in the first patch, but has not yet appeared in the second,
we would like to know the time scale on which copies of $B$ move between patches by migration
(supposing that no $B$ allele arises independently by mutation in the meantime).
To determine this, 
we study the fine-scale genealogy of alleles that transit between patches,
obtaining along the way other useful information about the genealogy of $B$ alleles. 
Doing this we arrive at equation \eqref{eqn:migrate} for the rate at which 
an allele established in one patch spreads to a neighboring one by migration.

@@

To understand the process of adaptation by migration,
we therefore need to look more closely at the rare families of $B$ alleles 
far from the original patch.
It is reasonable to model each such family as a branching process,
i.e., assuming that each individual
migrates, then reproduces, independently of all other family members.
Migration takes the individual in a random direction
for a random distance whose mean square is $\sigma^2$,
and if the individual arrives to a patch where the $B$ allele is favored, 
we drop the individual from the family.
Since the $B$ allele is uniformly deleterious elsewhere,
we can decouple reproduction and migration
by first forming a tree in which each individual has a random number of offspring,
then assigning spatial locations to each individual in the tree,
and finally pruning any branches that wander into a patch.
This pruning removes both branches leading back into the already established patch (which are irrelevant)
and branches leading into the new patch (we will need to count these).
Since we are concerned with the rare migrant families that transit quickly between patches, 
we ignore back migrants to the already established patch. 
% will contribute little, since they are going the wrong direction;
% so to simplify the analysis we ignore this detail.  % but could add this detail in an appendix

@@

This suggests that families can be well-described by motion of a central ``trunk'',
which dies at a random time whose distribution is given by $k_e(t)$,
surrounded by ``branches'',
% and conditional on survival of this trunk, the family is composed of side ``branches'' off of this trunk,
as depicted in Figure~\ref{fig:branching_decomp}.
We can understand this with a decomposition described by \citet{geiger1999elementary}.
% (see also \citet{chauvin1991growing}). 
% (lyons1995conceptual deals with conditioned on eventual nonextinction)
When we condition on survival until $t$, we condition on survival of at least one trunk lineage
(the long red lineage in Figure \ref{fig:branching_decomp}).
Given this lineage, the other reproduction events are undistinguished --
% each individual along the trunk lineage must give birth to at least one offspring that survives until time $t$,
% and all other reproduction events are unconditioned.
% In other words, 
the genealogy of a family that survives until $t$
can be decomposed into a trunk that lasts until $t$
and that sprouts independent, unconditioned branches that may or may not survive until $t$.
Concretely, if we write $Z_t$ for the number of offspring in the branching process alive at time $t$,
% (so $\P\{Z_t = 0\} = k_e(t)$),
then
\begin{align} \label{eqn:Zt_decomposition}
  \left( Z_t \; \vert \; Z_t>0 \right) \deq 1 + \sum_{s=0}^t \sum_{k=1}^{W_s-1} Z^{(s,k)}_{t-s},
\end{align}
where each $Z^{(s,k)}$ is an independent copy of $Z$ 
that arose from the main branch in generation $s$,
and each $W_s$ is independent with distribution 
%% equation (2.3) in Geiger:
%% $\P\{ W_s = k\} = \sum_{j=k}^\infty \P\{ W = j \} k_e(t-s-1)^{j-k} (1-k_e(t-s-1))/(1-k_e(t-s))$, 
%% and equation (2.5) in Geiger:
that tends to $\P\{W_s=k\} \simeq \P\{W\ge k\}/\E[W]$ as $s \to \infty$,
where $W$ is the unconditioned offspring number distribution.  \linelabel{ll:branching_process_fix}
In particular, $\E[W_s-1] \simeq \E[W(W-1)]/\E[W]$;
if $W$ is Poisson then this is equal to $e^{-s_m}$.
Since each subfamily $Z^{(s,k)}$ is a branching process with mean offspring number less than one,
most of these die out quickly,
and so $Z_t$ is composed of a cluster of recently split subfamilies.
If we take the expectation of \eqref{eqn:Zt_decomposition},
since $\E[Z_t] = e^{-s_m t}$,
we get the limiting mean family size conditioned on nonextinction:
\begin{align} \label{eqn:mean_K}
    \E[K] 
    = 1 + \sum_{s=0}^\infty \E[W_s-1] e^{-s_m t}
    \approx \E[W(W-1)] / ( \E[W] s_m  )
\end{align}

@@

\begin{figure}[ht!!]
  \begin{center}
    \includegraphics{branching-concept}
  \end{center}
  \caption{
      \textbf{(a)} 
      Cartoon of the decomposition described in the text. 
    Migrants families that get further than $r_0$ from the already-established patch (grey circle)
    are each composed of a ``trunk'' lineage (in red), whose spatial motion we follow,
    embellished by transient branches.
    For the depicted long-lived family, these are further distinguished by those still alive at time $t$ (in black)
    and those that did not survive until $t$ (in grey).
    Figure~\ref{fig:sim_snapshots} has a depiction of such a family, from a simulation. 
    Note that the spatial motions of lineages are independent random walks
    (and hence should be much wigglier than in this cartoon, which confounds space and time).
      \textbf{(b)} 
      Diagram of the area $A'$ used in equation \eqref{eqn:migrate}:
      $R$ is the shortest distance from the old patch to the new,
      and $A'$ is the area of the portion of the new patch that is no more than distance $R+\sigma/\sqrt{2s_m}$
      from the old patch.
} \label{fig:branching_decomp}
\end{figure}

@@

The remaining term is $q(S)$, the integral of equation \eqref{eqn:eqfreq} over the patch.
Since the patches are at distance $R$,
if the new patch has width smaller than $\sigma/\sqrt{2s_m}$, 
then $q(S)$ is approximately $q(R)$ multiplied by the area of the new patch.
If $S$ is larger, then we should use the area of the closest strip
of width $\sigma/\sqrt{2s_m}$ to the original patch;
see figure \ref{fig:branching_decomp}, and section \secref{apx:qS} for more details.
In either case, denote this area $A'$;
for numerical examples we compute $q(S)$ directly.

@@

To keep the expressions simple, in the remainder of this section we only compute quantities for one dimension.
By Bayes' rule, 
\begin{equation} \label{eqn:haplen_cdf}
    % \E[e^{-\ell \tau}|\tau<\infty]  =
    \P\{ L > \ell \}
    = 
    \exp\left\{{-\frac{R}{\sigma}\left(\sqrt{2(\ell+s_m)} - \sqrt{2s_m}\right)}\right\} .
\end{equation}
This can be differentiated to find that the expected transit time is
\begin{equation} 
  \E[\tau \given \tau<\infty ] = (R\sqrt{2s_m}/\sigma)\times 1/(2s_m) \label{eqn:mean_tau}
\end{equation}
and that $\var[\tau \given \tau<\infty ] = (R\sqrt{2s_m}/\sigma) \times 1/(2s_m)^2$.
(A saddle point approximation provides an alternative route to these expressions.) \revpoint{1}{4}

@@

Coat color in the rock pocket mouse
\emph{Chaetodipus intermedius} is a classic example of local
adaptation 
\citep{benson1933concealing,DiceBlossom1937}.
These mice live on rocky outcrops throughout the southwestern United States and northern Mexico, 
and generally have a light pelage similar to the predominant rock color.
However, in some regions these mice live on darker substrates (e.g., old lava flows),
and in these patches have much darker pigmentation, 
presumably to avoid visual predators.
Some of the largest patches of dark rock range from 10km to 100km wide
and lie 50--400km from each other,
and dark-colored populations of \emph{C.\ intermedius} have been found on many of these.
(However, patches of all sizes occur across all scales in a heterogeneous manner.)
\revpoint{2}{7}
\citet{nachman2003genetic} demonstrated that on one of these flows (Pinacate), 
much of the change in coloration is due to an allele at the MC1R locus.
This dark allele differs from the light allele by four amino acid changes, 
and has a dominant or partially dominant effect depending on the measure of coat color.
The Pinacate allele is not present in a number of other populations with dark pelage,
suggesting these populations have adapted in parallel \citep{nachman2003genetic,hoekstra2003different}.
However, \citet{hoekstra2004local} reasoned that, elsewhere in the range, 
multiple close dark outcrops may share a dark phenotype whose genetic basis has been spread by migration
despite intervening light habitat.

@@

In the development above
we need to approximate the integral of equation \eqref{eqn:eqfreq}
across the area occupied by the new patch,
$q(S) = \int_S q(x) dx$.
The precise answer depends on the shape and orientation of the patches;
but we aim for a usable approximation,
primarily in terms of the shortest distance from the old patch to the new patch, denoted $R$,
and assuming $R$ is large enough we can take equation \eqref{eqn:eqfreq} as an equality.
